---
title: "02_Uncertainty: week 2"
output: html_document
date: "2024-12-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setting the data and working directory:
```{r dataset}
setwd("/Users/carmencilla/Desktop/Master EUR/Block 2/Data Science and HR Analytics/FEM11213-Coronas/")
browser <- read.csv("Data/web-browsers.csv")
dim(browser)
head(browser)
```
## BH Algorithm 
In this lecture, BH Algorithm.
Multi-hypothesis testing. Pick a target value for q (0.1). Obtain p-values, re-order p and compute largest i such that: 
(p(i) =< i/m * q) = k. Reject H(i) for all i up to k. 

Note:this procedure substitutes signficance level!! So no need to check for significance. 

## Example 1: browse dataset

```{r BH Algorithm}
spendy <- glm(log(spend) ~ . -id, data=browser)
  round(summary(spendy)$coef,2)
```
Comments:  
All variables significant except controls for race.  
Issue with this method: sometimes slope too close to zero: therefore, sometimes useless procedure... 


```{r }
pval <- summary(spendy)$coef[-1, "Pr(>|t|)"]
  pvalrank <- rank(pval)
  reject <- ifelse(pval< (0.1/9)*pvalrank, 2, 1) 
  
  print(pval)
  print(reject)
```


## Example 2: semiconductor dataset
Context of the example: we wanna look if the chip fails. 

```{r data for example 2}
SC <- read.csv("../Data/semiconductor.csv")
dim(SC)
head(SC)
```

SC: More columns (201)

```{r regression}
full_sc <- glm(FAIL ~ ., data=SC, family=binomial)
pvals_sc <- summary(full)$coef[-1,4] #-1 to drop the intercept
print(pvals_sc)
```

```{r histogram for pvalues}
hist(pvals_sc, xlab="p-value", main="", col="lightblue")
```

Pvalues ranged between 0-1. 

```{r setting the cut}
fdr_cut <- function(pvals_sc, q=0.1){ ## The FDR threshold (e.g., 0.1 means a 10% FDR is allowed).
  pvals_sc <- sort(pvals_sc[!is.na(pvals_sc)]) #Sort and Filter P-Values: essential for the BH algorithm!!
  N <- length(pvals_sc) #Total number of p-values after removing NA.
  k <- rank(pvals_sc, ties.method="min") #Ranks of the p-values (from smallest to largest). Assigns the smallest rank to tied values.
  alpha_sc <- max(pvals_sc[ pvals_sc<= (q*k/(N+1)) ]) #alpha_sc = the largest p-value below the BH-adjusted threshold. BH condition!! 
  
  plot(pvals_sc, log="xy", xlab="order", main=sprintf("FDR of %g",q),  ## log-log plot to visualize results!!
   ylab="p-value", bty="n", col=c(8,2)[(pvals_sc<=alpha_sc) + 1], pch=20)
  lines(1:N, q*(1:N)/(N+1)) ## Line representing the BH threshold

  return(alpha_sc)
}

fdr_cut(pvals_sc)
```
## Interpreting the graph: 
  Red dots interpretation: we aim to have a simpler model (regularization). We use this regression to remove coefficients. Simple way of selection. 
  In the histogram, we saw many coefficients with pvalues being 0!! This means many significant coefficients!!
  Binomial model. 

## Understanding the difference between pvalue and FDR:
  - PValue: we aree sure that every individual observations has a 5% chance of being false positive. This is, 5% limit
  to reject H0. 5% chance of have Type I Error. More trials, significant increase in FP. 
  - FDR: controls for the FP among the significant results. We allow for a 10% of results on average neing false
  positive. FDR does not account for individual chances of FP. 
  
  This means that, if we have 100 and 20 of them are significant, 10% of them are potential FP (2 out of 20). 

```{r experiment BH}  
full_sc2 <- glm(FAIL ~ ., data=SC, family=binomial)
pvals_sc2 <- summary(full)$coef[-1,4] #-1 to drop the intercept
print(pvals_sc2)
```

```{r setting the cut 2}
fdr_cut2 <- function(pvals_sc2, q=0.01){ ## The FDR threshold (e.g., 0.1 means a 10% FDR is allowed).
  pvals_sc2 <- sort(pvals_sc2[!is.na(pvals_sc2)]) #Sort and Filter P-Values: essential for the BH algorithm!!
  N2 <- length(pvals_sc2) #Total number of p-values after removing NA.
  k2 <- rank(pvals_sc2, ties.method="min") #Ranks of the p-values (from smallest to largest). Assigns the smallest rank to tied values.
  alpha_sc2 <- max(pvals_sc2[ pvals_sc2<= (q*k2/(N2+1)) ]) #alpha_sc = the largest p-value below the BH-adjusted threshold. BH condition!! 
  
  plot(pvals_sc2, log="xy", xlab="order", main=sprintf("FDR of %g",q),  ## log-log plot to visualize results!!
   ylab="p-value", bty="n", col=c(8,2)[(pvals_s2c<=alpha_sc2) + 1], pch=20)
  lines(1:N2, q*(1:N2)/(N2+1)) ## Line representing the BH threshold

  return(alpha_sc2)
}

fdr_cut2(pvals_sc2)
```

Lowering the q, we diminish the threshold to accept non-zero coefficients, therefore accepting less coefficients. 